High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-08-23 paper/main.tex
working on methods section-- low-dimensional representations of dynanmic correlations

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
correlations between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}

\textbf{JRM STOPPED HERE}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{weight function} that takes a time $t$ as input,
and returns how much the observed data every timepoint (including $t$)
contribute to the correlations at time $t$ (Fig.~\ref{fig:weights}).
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of time-varying weights.  Each panel
      displays per-timepoint weights at $t = 50$, evaluated for 100
      timepoints ($1, ..., 100$).  a. Uniform weights.} The weights
    are timepoint-invariant; observations at all timepoints are
    weighted equally, and do not change as a function of $t$.  This is
    a special case of weight function that reduces dynamic
    correlations to static correlations.  \textbf{b. Dirac delta
      function.} Only the observation at timepoint $t$ is given weight
    (of 1), and weights for observations at all other timepoints are
    set to 0.  \textbf{c. Gaussian weights.} Each observation's
    weights fall off in time according to a Gaussian probability
    density function centered on $\mu = t$.  Weights derived using
    several different example variance parameters ($\sigma^2$ are
    displayed.  \textbf{d. Laplace weights.}  Each observation's
    weights fall off in time according to a Laplace probability
    density function centered on $\mu = t$.  Weights derived using
    several different example scale parameters ($b$) are displayed.
    \textbf{e. Mexican hat (Ricker wavelet) weights.}  Each
    observation's weights fall off in time according to a Ricker
    wavelet centered on $t$.  This function highlights the
    \textit{contrasts} between local versus surrounding activity
    patterns in estimating dynamic correlations. Weights derived using
    several different example width parameters ($\sigma$) are
    displayed.  }
  \label{fig:weights}
\end{figure}

Given a weight function $w(t)$ for timepoint $t$, evaluated at
timepoints in the interval $\left[ 1, ..., T \right]$, we can extend the static correlation formula
in Equation~\ref{eqn:corr} to reflect an \textit{instantaneous
  correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}(\mathbf{X}_i, \mathbf{X}_j, t) &= \frac{\sum_{t=1}^T \left(\mathbf{X}_i(t)
                                              -
                                              \widetilde{\mathbf{X}}_i(t)\right) \left(\mathbf{X}_j(t)
                                              -
                                              \widetilde{\mathbf{X}}_j(t)\right)}{\sqrt{\sum_{t=1}^T
                                              \widetilde{\sigma}^2_{\mathbf{X}_i}(t) 
                                              \widetilde{\sigma}^2_{\mathbf{X}_j}}(t)},~\mathrm{where}\\\label{eqn:timecorr}
  \widetilde{\mathbf{X}}_k(t) &= \sum_{i=1}^T
                       w(t, i)\mathbf{X}_k(i),\\
  \widetilde{\sigma}^2_{\mathbf{X}_k}(t) &= \sum_{i=1}^T \left( \mathbf{X}_k(i) -
                            \widetilde{\mathbf{X}}_k(t) \right)^2,
\end{align}
and $w(t, i)$ is shorthand for $w(t)$ evaluated at timepoint $i$.
Equation~\ref{eqn:timecorr} may be used to estimate the instantaneous
correlations between every pair of observations, at each timepoint
(i.e., $\mathbf{Y}$).

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}$.  Suppose that one
has access to a set of multiple observation matrices that reflect the
same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_1$,
$\mathbf{X}_2$, ..., $\mathbf{X}_P$ reflect the $T$ by $F$ observation
matrices for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{i=1}^P
  Z\left(Y_i(t)\right)^\mathrm{T} + Z\left(Y_i,(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_i(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_i$ and each
column of the average observations from all \textit{other}
participants, $\bar{\mathbf{X}}_{ \backslash i}$:
\begin{align}
  \bar{\mathbf{X}}_{ \backslash i} = R\left(\frac{1}{P-1}\sum_{i \in
  \backslash i} Z\left( \mathbf{X}_i \right) \right),
\end{align}
where $ \backslash i$ denotes the set of all participants other than
participant $i$. In this way, the $T$ by $\left( \frac{F^2 - F}{2}
  + F \right)$
matrix $\bar{\mathbf{C}}$ is the time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

\subsubsection*{Dimensionality reduction-based approaches to computing
  $\hat{\mathbf{Y}}_n$}

\subsubsection*{Graph theory-based approaches to computing
  $\hat{\mathbf{Y}}_n$}

\subsection*{Evaluation metrics}
We evaluate our approach to extracting dynamic correlations and
higher-order correlations using several metrics detailed next.  First,
we generated synthetic data using known time-varying correlations, and
then we evaluated the fidelity with which Equation~\ref{eqn:timecorr}
could recover those correlations (for synthetic datasets with
different properties, and using different kernels to define the
weights; Fig.~\ref{fig:weights}).  We then turned to a series of
analyses on a (real) neuroimaging dataset where the ground truth
correlations were \textit{not} known.  We evaluated whether the
recovered correlations could be used to accurately label held-out
neuroimaging data with the time at which it was collected.  We used this latter evaluations (using timepoint
decoding) as a proxy for
gauging how much explanatory power the recovered correlations held
with respect to the observed data.

@@

We applied timecorr, using delta and gaussian (width = 10) kernels
Fig.~\ref{fig:weights}) to each of these 
synthetic datasets, then correlated each recovered
correlation matrix with the ground truth.  We repeated this process 10
times and explored how recovery varies
with the kernel and the specific structure of the data. For the
ramping synthetic dataset (Fig.~\ref{fig:synthetic_data},  c.)  and for the
block synthetic dataset (Fig.~\ref{fig:synthetic_data},  d.)  we made further
comparisons of the timecorr recovered correlation matrices. We
compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  c.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
c.) from Equation~\ref{eqn:ramping}. We also compared the block recovered correlation matrices in to
the block specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  d.).

@@

To assess the performance of dynamic correlation recovery using
timecorr, we varied width the kernel and the specific structure of the
data. We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of the following  
synthetic datasets: constant, random, ramping, and block.  We then correlated each recovered
correlation matrix with the ground truth.

>>>>

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$ (Fig.~\ref{fig:kernels}).
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example variance parameters ($\sigma^2$ are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example scale parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\\\label{eqn:timecorr}
  \widetilde{\mathbf{X}}_{\kappa_t}(t, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here $\mathrm{timecorr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j),
  \kappa_t)$ reflects the correlation at time $t$ between columns $i$
  and $j$ of $\mathbf{X}$, estimated using the kernel $\kappa_t$.

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(Y_n^p(t)\right)^\mathrm{T} + Z\left(Y_n^p(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_n^p(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = R\left(\frac{1}{P-1}\sum_{q \in
  \backslash p} Z\left( \mathbf{X}_n^q \right) \right),
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\mathcal{O}(K^2)$
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

\subsection*{Low-dimensional representations of dynamic
  correlations}

Given a $T$ by $\mathcal{O}(K^2)$ matrix of dynamic correlations,
$\mathbf{Y}_n$, we propose two general approaches to computing a $T$ by $K$
low-dimensional representation of these correlations, $\mathbf{X}_n$.
The first approach uses dimensionality reduction algorithms to project
$\mathbf{Y}_n$ onto a $K$-dimensional space.  The second approach uses
graph-theoretic measures to characterize the relative positions of
each feature ($k \in \left[1, ..., K \right]$) in the network defined
by the correlation matrix at each timepoint. \textbf{JRM STOPPED HERE}

\subsubsection*{Dimensionality reduction-based approaches to computing
  $\mathbf{X}_n$}

\subsubsection*{Graph theoretic approaches to computing
  $\mathbf{X}_n$}

@@

\subsection*{Evaluation metrics}
We evaluate our approach to extracting dynamic correlations and
higher-order correlations using several metrics detailed next.  First,
we generated synthetic data using known time-varying correlations, and
then we evaluated the fidelity with which Equation~\ref{eqn:timecorr}
could recover those correlations (for synthetic datasets with
different properties, and using different kernels to define the
weights; Fig.~\ref{fig:kernels}).  We then turned to a series of
analyses on a (real) neuroimaging dataset where the ground truth
correlations were \textit{not} known.  We evaluated whether the
recovered correlations could be used to accurately label held-out
neuroimaging data with the time at which it was collected.  We used this latter evaluations (using timepoint
decoding) as a proxy for
gauging how much explanatory power the recovered correlations held
with respect to the observed data.

@@

We applied timecorr, using delta and gaussian (width = 10) kernels
Fig.~\ref{fig:kernels}) to each of these 
synthetic datasets, then correlated each recovered
correlation matrix with the ground truth.  We repeated this process 10
times and explored how recovery varies
with the kernel and the specific structure of the data. For the
ramping synthetic dataset (Fig.~\ref{fig:synthetic_data},  c.)  and for the
block synthetic dataset (Fig.~\ref{fig:synthetic_data},  d.)  we made further
comparisons of the timecorr recovered correlation matrices. We
compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  c.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
c.) from Equation~\ref{eqn:ramping}. We also compared the block recovered correlation matrices in to
the block specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  d.).

@@

To assess the performance of dynamic correlation recovery using
timecorr, we varied width the kernel and the specific structure of the
data. We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:kernels}) to each of the following  
synthetic datasets: constant, random, ramping, and block.  We then correlated each recovered
correlation matrix with the ground truth.
